Gradual Soundness: Lessons from Static Python

### 2022-01-26 pj-7.1/paper.tex
pj: many sp edits sec 2 3 5

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{} and related proposals~\cite{pep484}.
Its static semantics for types is inspired by the Pyre~\cite{pyre}, mypy~\cite{mypy}, and Pyright~\cite{pyright}
optional type checkers.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type is available as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden on its target users.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no plans at present to support all \PEP{} types.
The goal is to implement a small language of types efficiently.

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Union types are well-supported statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),
but are, for the most part, not supported at run-time~(\cref{s:eval-types}).
The only exception is that binary unions of a type and the Python \code{None}
value get enforced with a run-time check ($\sptunion{\sptype, \sptnone}$, or
$\sptoptional{\sptype}$).

@@

%% 2021-12-14 TODO:
%% - anything to say about Any being a "meta" type in Python typing module?
%%   2022-01-10 bg: not sure what the consequences of "meta type" are

Like most gradual languages, \SP{} includes a dynamic type
that allows untyped expressions within a statically-typed context.
Whenever an expression or variable lacks a type annotation, \SP{} uses the
dynamic type as a default and skips most \emph{but not all\/} static checks.
Programmers can thus add annotations to part of a module
while leaving the rest unannotated.

%% TODO avoid typo ... shorten?
In addition to serving as a default, the dynamic type also replaces any undefined types.
For example, suppose that the type $\sptflout{}$ is undefined in \SP{}
(i.e., it is a misspelling of $\sptfloat$) and that a
program contains the annotation $\sptchkdict{\sptstr}{\sptflout}$, which describes a
dictionary with integer keys and $\sptflout$-typed values.
\SP{} interprets this annotation as $\sptchkdict{\sptint}{\sptdyn}$ and allows any sort
of values inside the dictionary at runtime.

TODO reword.
At first glance, this behavior might seem odd: why not reject undefined types with an error?
The reason is to reduce friction with existing \PEP{} annotations.
Pyre is a mature static type system that can check more types than \SP{} knows how to enforce
efficiently, e.g., the type $\sptset{\sptint}$ for sets of integers.
By replacing such types with dynamic, \SP{} can accept Pyre code that it does not yet fully understand.

The \SP{} dynamic type is also different than the flexible dynamic type provided
by gradual languages that satisfy the gradual
guarantees~\cite{svcb-snapl-2015}.
In \SP{}, replacing part of a type with dynamic can lead to a static error when
the compiler analyzes that part of the module.
Similarly, the addition of a dynamic type can lead to a run-time error because of the
strict checks that \SP{} applies to generic types.
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
%% \footnote{Technically, \SP{} does not even
%% have syntax for the dynamic type. The idiomatic way to ask for dynamic is to
%% omit a type annotation.}
%% \footnote{We
%% assume a standard type precision relation. Of course, one could argue that these programs
%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
%%   %% tailored = describes exactly the migrations that SP allows
%% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m} method to return the dynamic type.

  \item
    \Cref{f:gg-failure-dyn} sends a checked dictionary with integer values to a
    function that expects dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These behaviors enable efficient run-time checks and additional type-directed
optimizations.
Part of the reason \SP{} weighs these benefits more heavily than the costs
is that programmers have to opt-in to a feature in order to risk the errors.
The static error above comes only after enabling the \SP{} compiler on a module.
The dynamic error comes only after adopting a concrete type (for checked dictionaries).

More to the point, \SP{} has little interest in guarantees about
\emph{removing an arbitrary annotation}.
It is, however, very interested in making sure that \emph{an untyped
module compiles with minimal code changes}.

@@

Parameterized types (generics) are more difficult to enforce.
For example, suppose that a typed function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for dictionaries is not enough to protect the typed
function because it fails to validate the elements inside a dictionary.
The \SP{} solution is to provide two incompatible versions of every parameterized type,
both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
and concrete versions at present.
In particular, user-defined classes do not have concrete support.
The shallow check for a user-defined class, however, guarantees the types of all
non-generic fields and methods.}
For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores the elements inside a dict.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    is merely a promise of dict values at run-time; it does not promise anything about
    dict keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
    keys and values, but is incompatible with built-in Python dictionaries.
    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

The upside of concrete types is that they enforce stronger type constraints.
These constraints can catch bugs and usually lead to more-efficient code.
Faster performance is not guaranteed, though, because it depends on how the dict
gets used.
\Cref{t:shallow-vs-concrete} presents a detailed comparison.
Shallow types pay a constant-time cost whenever an untyped value enters typed code
and whenever a typed context reads from a shallow value.
The net cost of reads may be high.
Concrete types pay a potentially-high cost for constructors,
a constant-time cost for casts and for writes,
and zero cost for dereferences.
If a value is frequently written to and rarely read from, shallow types may actually
be faster.

\begin{table}[t]
  \caption{Worst-case costs for shallow and concrete types}
  \label{t:shallow-vs-concrete}

  \centering
  \(\begin{array}{lrrrr}
     & \mbox{Constructor} & \mbox{Cast} & \mbox{Read} & \mbox{Write} \\\hline
    \sptrawpydict                &           0 & O(1) & O(1) &     0 \\
    \sptchkdict{K}{V} &        O(N) & O(1) &    0 &  O(1)
  \end{array}\)
\end{table}

@@

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
Each typed class can furthermore assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
One extra step arises when an untyped class overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

The \SP{} runtime system, formally known as Cinder, extends
CPython 3.8 in several ways to take full advantage of static types.
Cinder includes tailored bytecode instructions, virtual method tables,
concrete datatypes, a registry of typed modules, and a JIT compiler.

These main ingredients of Cinder may be of interest to other teams seeking to
add sound gradual typing to an existing language.
For example, Node developers may wish to fork that runtime and experiment with
bytecode instructions that leverage sound static types.

\subsection{Bytecode Instructions}

@@

\subsection{Virtual Method Tables}
\label{s:vtable}

@@

The implementation of vtables happens to be built on the Python \code{vectorcall}
protocol; it is not a from-scratch development.
Additionally, however, Cinder vtables are specialized to check untyped overrides
of typed methods against their expected return type (instead of using a layer of
indirection for the check, as found languages that enforce types with higher-order
wrappers~\cite{clzv-ecoop-2018,gf-icfp-2018}).

%% beware: with vtables, method resolution is different than Python b/c args. get
%%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)

%% > This difference in behavior isn’t so much desired as just a consequence of making
%% > `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
%% > the callable is placed on the stack, then the arguments, then there is a
%% > `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
%% > means that first the callable is resolved, then the arguments. But with
%% > `INVOKE_METHOD` we want to resolve the callable as part of the invoke
%% > itself, since this gives us opportunity to inline-cache the target of the
%% > call instead of always having to call something dynamic and unknown that’s
%% > on the stack. So that necessarily implies that first the arguments are
%% > resolved and placed on the stack, then the callable is resolved as part of
%% > the `INVOKE_METHOD`. -Carl

\subsection{Concrete Data Structures}
\label{s:chkdict-impl}

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All write operations check their input.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
    and $V$ is equal to $V'$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

\subsection{Classloader}
\label{s:classloader}

\SP{} keeps track of typed functions and typed classes with a specialized classloader.
At runtime, the classloader keeps a registry of typed objects and whether or not they
are mutable.
The registry helps the bytecode reference objects and types via their module names.
The mutability bit lets \SP{} optimize access for immutable data while soundly allowing
patches to mutable data.

\subsection{Method-at-a-Time JIT}
\label{s:jit}

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~\cref{s:c-types}, module-level variables,
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

%% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
There are two cases in which \SP{} types impose constraints on untyped Python code.
First, concrete generic types (e.g., $\sptrawchkdict$)
reject inputs that were not created via a checked contructor.
If a client of a Python module decides to impose concrete types, the Python module
may need to change.
Second, typed modules prevent updates to module-level variables.

Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs.
Similarly, concrete types such as $\sptrawchkdict$ also
check their writes.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness

@@

\SP{} does not ascribe types to the following Python features.
These are not covered in our model because the implementation simply
assigns the dynamic type and lets the runtime treat them as untyped
Python code.
For each, the dynamic type is a very reasonable choice;
accurate static types would be burdensome to maintain.

@@

\SP{} is one of many type systems for Python.
Other optional type checkers include Pyre~\cite{pyre}, mypy~\cite{mypy}, and Pyright~\cite{pyright};
these systems do not ensure safe interactions with untyped Python.
Another sound type system is Reticulated,
which enforces structural types (rather than nominal) using its transient
semantics~\cite{vss-popl-2017,v-thesis-2019}.
Other \emph{optimizing} type systems for Python
include Reticulated and mypyc.
The latter compiles Python source to C extension modules; unlike \SP{},
it does not offer a pure-Python developer experience.

>>>>

The \SP{} type system is a unique synthesis of ideas from the gradual typing literature
and prior work on types for Python.
Its syntax for types is based on the definitions in \PEP{} and related proposals~\cite{pep484}.
Its static semantics for types is inspired by the Pyre~\cite{pyre} and mypy~\cite{mypy}
optional type checkers.
And its strategy for run-time checks is adapted from Nom~\cite{mt-oopsla-2017,mt-oopsla-2021},
a research language with compelling performance results.
\SP{}'s novelty comes from the following engineering constraints:
\begin{itemize}
  \item
    Performance is the bottom line.
    At the end of the day, \SP{} needs to make code run faster in production.
  \item
    Critical code often relies on first-order functions and objects,
    at least in the Instagram web server.
    Precise types for higher-order functions, for first-class classes, and even for
    locally-defined functions and classes are thus a low priority; instead, the
    dynamic type is available as a coarse approximation.
  \item
    The codebase in which \SP{} is being applied makes heavy use of
    \PEP{} type annotations and is regularly analyzed by Pyre.
    \SP{} is therefore compatible with the \PEP{} syntax to reduce the
    adoption burden on its target users.
    %% depth-first approach? (the alternatives are not very clear ... depth vs breadth?)
  \item
    A module-level granularity is acceptable.
    Once \SP{} is enabled for a module, it compiles all code in that module
    including expressions that have the dynamic type~(\cref{s:type-dynamic}).
    If this behavior is a problem for certain expressions, it is easy enough to move
    them to an untyped Python module.
\end{itemize}
%
In general, the types that \SP{} implements all enable significant optimizations
and can be validated quickly with casts.
Types that do not meet these criteria are deferred to the mature optional
type checkers that already exist for Python.
In fact, \SP{} has no plans at present to support all \PEP{} types.
The goal is to implement a small language of types efficiently.

The core supported types describe basic data (integers, booleans, strings),
data structures (lists, dicts, promises),
and nominal classes.
Union types are tracked statically and narrow down in accordance
with type tests (a form of occurrence typing~\cite{tf-icfp-2010,gsk-esop-2011}),\footnote{
  For example, if \code{x} is an integer and \code{y} is a string, then the
  expression \code{x or y} has a union type.}
but are not generally supported at run-time~(\cref{s:eval-types}).
The only exception is that binary unions of a type and the Python \code{None}
value get enforced with a run-time check ($\sptunion{\sptype, \sptnone}$, or
$\sptoptional{\sptype}$).

@@

\SP{} includes a dynamic type that allows untyped expressions within a
statically-typed context.
Whenever an expression or variable lacks a type annotation, \SP{} uses the
dynamic type as a default and skips most static checks.
The overall goal of this design is to let programmers add simple annotations to
part of a Python module while leaving the rest untyped.

In addition to unannotated positions, the dynamic type is also the default for
\PEP{} types that \SP{} does not yet understand.
If existing code uses the type $\sptset{\sptint}$ for sets of integers,
then \SP{} replaces it with dynamic.
Thus, \SP{} focuses on types that it can soundly optimize without forcing
programmers to remove types that other systems (e.g., mypy) can check.

The behavior of the \SP{} dynamic type is subtle, however, because dynamic code
faces more restrictions than untyped Python code.
Replacing part of a type with dynamic can lead to either a static error or
a run-time error.
\Cref{fig:gg-failure} presents two examples, one for each kind of failure:
%% \footnote{Technically, \SP{} does not even
%% have syntax for the dynamic type. The idiomatic way to ask for dynamic is to
%% omit a type annotation.}
%% \footnote{We
%% assume a standard type precision relation. Of course, one could argue that these programs
%% do not break the gradual guarantees for a type precision relation that is tailored to \SP{}.}
%%   %% tailored = describes exactly the migrations that SP allows
%% \footnote{In addition, C types~(\cref{s:c-types}) are incompatible with type dynamic.}
\begin{itemize}
  \item
    \Cref{f:gg-failure-stat} presents a fully-typed class and a partially-typed subclass.
    The subclass definition raises a compile-time error because it attempts to override
    the typed \code{m} method to return the dynamic type.

  \item
    \Cref{f:gg-failure-dyn} sends a checked dictionary with integer values to a
    function that expects dictionaries with dynamic values.
    At runtime, the function rejects this argument because its type is not an
    exact match.
\end{itemize}
These extra restrictions make the \SP{} dynamic type different than the flexible
dynamic type provided by languages that satisfy the gradual guarantees~\cite{svcb-snapl-2015}.
But they also enable efficient run-time checks for generics and type-directed optimizations
for method calls.
Part of the reason \SP{} weighs these benefits more heavily than the gradual guarantees
is that programmers have to opt-in to a feature in order to risk the errors.
The static error above comes only after enabling the \SP{} compiler on both a parent
class and its subclass.
The dynamic error comes only after adopting a concrete type for checked dictionaries.
More to the point, \SP{} is less interested in guarantees about \emph{removing
an arbitrary annotation} and more interested in making sure that \emph{an
untyped module compiles with minimal code changes}.

@@

Parameterized types (generics) are more difficult to enforce.
For example, suppose that a typed function expects inputs that match
the type \code{Dict[int,int]}; that is, dictionaries with integer keys and
integer values.
A constant-time check for dictionaries is not enough to protect the typed
function because it fails to validate the elements inside a dictionary.
The \SP{} solution is to provide two incompatible versions of every parameterized type,
both of which support efficient run-time checks.\footnote{Not all parameterized \SP{} types have both shallow
and concrete versions at present.
In particular, user-defined classes do not have concrete support.
The shallow check for a user-defined class, however, guarantees the types of all
non-generic fields and methods.}
For dictionaries, \SP{} provides the types $\sptrawpydict$ and $\sptrawchkdict$:
\begin{enumerate}
  \item
    The \emph{shallow} type $\sptrawpydict$ gets enforced with a constant-time check
    for dict values that ignores the elements inside a dict.
    Consequently, a parameterized surface type such as $\sptpydict{K}{V}$
    is merely a promise of dict values at run-time; it does not promise anything about
    keys and values.
  \item
    The \emph{concrete} type $\sptchkdict{K}{V}$ guarantees well-typed
    keys and values, but is incompatible with built-in Python dictionaries.
    Programmers must use a type-annotated contructor to make a concrete-typed checked dict.
\end{enumerate}
%
These options let programmers decide whether to edit their code to support concrete types
or leave it as-is with shallow types.
Often, the shallow types are more attractive because concrete types impose
nonlocal changes.
If one type changes to concrete, then several changes may need to happen:
all values that reach this type must be initialized as concrete,
all typed clients of these values must expect concrete values,
all values that reach those clients must be concrete,
and so on and so forth.

The upside of concrete types is that they enforce stronger type constraints.
These constraints can catch bugs and always lead to more-efficient typed code.
Faster performance is not guaranteed, though, because concrete types must
check all writes from untyped contexts.
\Cref{t:shallow-vs-concrete} presents in more detail the worst-case costs of
shallow and concrete dict types.
Shallow types pay a constant-time cost whenever an untyped value enters typed code
and whenever a typed context reads from a shallow value.
Concrete types pay a potentially-high cost for constructors,
a constant-time cost for casts and for writes in untyped code,
and zero cost for dereferences.
If a value travels to untyped code, gets written to many times, and gets read few times,
then the shallow types may run faster overall.

\begin{table}[t]
  \caption{Worst-case costs for shallow and concrete types (T = typed, U = untyped)}
  \label{t:shallow-vs-concrete}

  \centering
  \(\begin{array}{l@{}rr@{\qquad}rr@{~~}@{\qquad}r@{~~}r}
                   & \mbox{Constructor} & \mbox{Cast} & \mbox{T-Read} & \mbox{T-Write} & \mbox{U-Read} & \mbox{U-Write} \\\hline
    \sptrawpydict         &           0 &        O(1) &          O(1) &              0 &             0 &              0 \\
    \sptchkdict{K}{V}     &        O(N) &        O(1) &             0 &              0 &             0 &           O(1)
  \end{array}\)
\end{table}

@@

\end{enumerate}
%
With this context in mind, \SP{} keeps track of whether each class is typed or untyped.
Each typed class can furthermore assume that if a method has a precise (non-dynamic)
type, then all typed overrides are subtypes of this type.
\SP{} can therefore optimize dispatch from typed callers to typed methods
and use checked dispatch for other combinations.
One extra step arises when an untyped method overrides a typed method.
In this case, \SP{} creates a wrapper around the overriding method to check
that it computes type-correct results.
% Because these wrappers perform a simple first-order check, they
% are supported by an efficiently implementation that cooperates with
% \SP{}
The wrappers are handled efficiently by a tailored implementation of vtables~(\cref{s:vtable}).

@@

The \SP{} runtime system, formally known as Cinder, extends
CPython 3.8 in several ways to take full advantage of static types.
Cinder includes tailored bytecode instructions, virtual method tables,
concrete datatypes, a registry of typed modules, and a JIT compiler.
These main ingredients of Cinder may be of interest to other teams seeking to
add sound gradual typing to an existing language.
For example, Node developers may wish to fork V8 and experiment with
bytecode instructions that leverage sound static types.

\paragraph{Bytecode Instructions}

@@

\paragraph{Virtual Method Tables}
\label{s:vtable}

@@

The implementation of vtables happens to be built on the Python \code{vectorcall}
protocol; it is not a from-scratch development.
Cinder vtables are specialized to check untyped overrides of typed methods
using a wrapper implemented in C rather than in Python, to reduce performance costs.

%% beware: with vtables, method resolution is different than Python b/c args. get
%%   resolved before the receiver (A.m(B) goes "A->B" in Py. and "B->A" in SP)
%%
%% > This difference in behavior isn’t so much desired as just a consequence of making
%% > `INVOKE_METHOD` optimizable. With a normal dynamic `CALL_FUNCTION`, first
%% > the callable is placed on the stack, then the arguments, then there is a
%% > `CALL_FUNCTION` (with number of args in oparg) to perform the call. This
%% > means that first the callable is resolved, then the arguments. But with
%% > `INVOKE_METHOD` we want to resolve the callable as part of the invoke
%% > itself, since this gives us opportunity to inline-cache the target of the
%% > call instead of always having to call something dynamic and unknown that’s
%% > on the stack. So that necessarily implies that first the arguments are
%% > resolved and placed on the stack, then the callable is resolved as part of
%% > the `INVOKE_METHOD`. -Carl

\paragraph{Concrete Data Structures}
\label{s:chkdict-impl}

@@

For example, the type $\sptchkdict{K}{V}$ describes a concrete dictionary
with keys of type $K$ and values of type $V$.
The implementation has three main components:
\begin{itemize}
  \item
    The constructor requires two type parameters and a Python dictionary ($\sptrawpydict$).
    It checks that the elements of the dictionary match the key and value types.
  \item
    All untyped writes must be protected by casts.
    Every operation that mutates or extends a checked dictionary must
    validate any untyped arguments that it receives.
  \item
    The dictionary value stores a compound type tag 
    (e.g., one token representing a three-part type $\sptchkdict{K}{V}$)
    to support casts from the dynamic type.
    When a concrete dictionary enters typed code from an untyped context,
    its tag must match the expected type exactly.
    For example, type $\sptchkdict{K}{V}$ matches the type
    $\sptchkdict{K'}{V'}$ only if $K$ is equal to $K'$
    and $V$ is equal to $V'$.
\end{itemize}
%
In general, other checked datatypes have the same three components:
a constructor, checked update functions, and a tag.

\paragraph{Classloader}
\label{s:classloader}

\SP{} keeps track of typed functions and typed classes with a specialized classloader.
At runtime, the classloader keeps a registry of typed objects.
The registry helps the bytecode reference objects and types via their module names.

\paragraph{Method-at-a-Time JIT}
\label{s:jit}

@@

The model intentionally does not cover all of Python.
Some aspects of \SP{} are left out because they are straightforward to
handle.
These include primitive types~(\cref{s:c-types}), module-level variables,
and the boundary to untyped Python~(\cref{s:py-boundary}).
%% Aspects of \SP{} that do not pertain to soundness, such as the quality of error messages,
%% are also omitted.
Other Python features are left out because their \SP{} semantics is identical
to standard Python~(\cref{s:py-boundary}).
Our model also does not cover the optimizer~(\cref{s:optimize}) because
a formal account of type-directed optimization would require substantial additional
components: a model of the \SP{} bytecode and a faithful rendering of its
transformations.

@@

%% TODO FILL only 2 cases? revisit after adding "changes to python" section in 2
There are two cases in which \SP{} types impose constraints on untyped Python code.
First, concrete generic types (e.g., $\sptrawchkdict$)
reject inputs that were not created via a checked constructor.
If a client of a Python module decides to impose concrete types, the Python module
may need to change.
Second, typed modules prevent updates to module-level variables.

Interactions with {open-world}~\cite{vss-popl-2017} Python code pose a minor
threat to soundness because
typed functions (and methods) cannot assume that their arguments are
well-typed.
All arguments sent from typed contexts get validated either statically or
via materialization casts, but arguments from untyped contexts are unchecked.
For this reason, \SP{} compiles every typed function to check its inputs;
the optimizer can bypass these checks for typed-to-typed calls~(\cref{s:optimize}).
Similarly, concrete types such as $\sptrawchkdict$ must check writes that
come from untyped contexts.
%% 2022-01-09 worth mentioning primitive types? they get cast like anything else (they don't materiazize)
%% 2022-01-13 don't think so --- leave prims alone, we don't want to frighten reviews with unsoundness

@@

\SP{} does not ascribe types to the following Python features.
These are not covered in the model because the implementation simply
assigns the dynamic type and lets the runtime treat them as untyped
Python code.
For each, the dynamic type is a reasonable choice;
accurate static types would be burdensome to maintain.

@@

\SP{} is one of many type systems for Python.
Other \emph{optional\/} type checkers include Pyre~\cite{pyre},
mypy~\cite{mypy}, PyType~\cite{pytype}, and Pyright~\cite{pyright}; these
systems do not ensure safe interactions with untyped Python.
Another sound type system is Reticulated,
which enforces structural types (rather than nominal) using its transient
semantics~\cite{vss-popl-2017,v-thesis-2019}.
Other \emph{optimizing} type systems for Python
include Reticulated and mypyc.
The latter compiles Python source to C extension modules; unlike \SP{},
it does not offer a pure-Python developer experience.
